Left-linear rules in adhesive categories

### 2024-02-12 icalp/fusione.tex
intro

As an initial step in performing this exploration, we observe that for left-linear
rules a number of fundamental properties that play a role in the
corresponding theory of rewriting need to be deeply revised, or plainly do not hold.

Firstly, Church-Rosser theorem does not work out of the box. In
particular, sequential independence between rewriting steps~\cite{}
expressed in terms of the existence of a so-called independence pair
is a canonical and well-understood notion of independence in the
linear case. \todo{P: Too technical? Forse qui e' piu' esplicativo un
  esempio} Whenever two steps are sequential independent
\begin{itemize}
\item they can be switched and 
\item the result of the switch is uniquely determined.
\end{itemize}

Here we observe that, in general, both property can fail: sequential
independent steps could be not switchable and different independent
pairs could exist leading to different switches of the same rewriting
steps.  A refined notion of independence needs to be considered to ensure
that independent rewriting steps can be switched, which is based in the concept
of fillers. The problem was identified
in~\cite{baldan2011adhesivity}, where a subclass of adhesive
categories was identified where independence as induced by a filler
coincides with sequential independence (as induced by an independence
pair). Here it is extended to $\mathcal{M}$-adhesive categories.

We face the problem in the setting of $\mathcal{M}$-adhesive
categories (which capture adhesive and quasi-adhesive categories).
A difficult task, since the most natural idea, put forward in~\cite{baldan2017domains}, 
of limiting sequential independence to those cases in which the switch is possible
and uniquely determined does fall short.
%
In this work we focus on a class of rewriting systems, which we call
\emph{very tame rewriting systems}, in $\mathcal{M}$-adhesive
categories where sequential independence ensures existence and
uniqueness of the switch and argue that this is the right setting for
dealing with left-linear rewriting systems.

@@

On the other hand, we show that an appropriate theory of rewriting can
be developed for very tame rewriting systems, recovering, sometimes in
weakened form, results which hold in the linear case.  Local
Curch-Rosser is fully recovered for very tame rewriting systems.

For linear rewriting systems sequential independence of derivations
can be also characterised in terms of the existence of a suitable
isomorphism between the colimit of the derivations, called
consistent permutations (which are also strictly related to the notion
of process). This result, which is folklore for categories of graphs,
is explicitly proved here for $\mathcal{M}$-adhesive categories. Based on such result one can deduce two fundamental consequences:
\begin{enumerate}
\item when switching a derivation consisting of several steps, the result only depends on the overall permutation, not on the order of the switches;
\item independence of rewriting steps is a ``global property'', independent of the position in which these steps are in the derivation.
\end{enumerate}

We show that for left-linear rewriting systems the characterisation of
shift equivalence based on consistent permutations fails: the
existence of a consistent permutation is only necessary but not
sufficient, hence it cannot be used for proving (1) and (2). However property (1) still holds. Instead only a weak form of (2) can be proved. This is intimately related to the possibility of expressing disnjuctive forms of causality.

DA RIUSARE\\
The characterisation of switch equivalence based on consistent permutations is the key to
derive some simple but fundamental properties for linear rewriting systems, namely
\begin{enumerate}
\item \emph{confluence of switching}: when transforming derivation
  sequences into equivalent ones by switching independent steps, the
  result not depends on the order in which switchings are applied, but
  only on the resulting permutation;

In particular, the two properties allows one to obtain a \emph{canonical form} for switching sequences: whenever two derivation sequences are switch equivalence via some permutation $\sigma$, then each one can be obtained from the other by applying switchings to pairs of rule applications which $\sigma$ reverses (the so-called inversions).

The failure of the characterisation of switch equivalence via
consistent permutations for left-linear rules leads to the
question asking whether the properties above hold.

>>>>

As an initial step in performing this exploration, we observe that for left-linear
rules a number of fundamental properties that play a role in the
corresponding theory of rewriting need to be deeply revised, or plainly do not hold.
%
More to the point, for linear rewriting systems sequential independence of 
two sequences of steps can be characterised in terms of the existence of 
what is called a consistent permutation among them, which can be obtained through 
an isomorphism between their colimits~\cite{xxx}. This result, which is folklore for 
various categories of graphs, is explicitly proved here for $\mathcal{M}$-adhesive categories. 
%
The characterisation of switch equivalence based on consistent permutations is the key to
derive some simple but fundamental properties for linear rewriting systems, namely
\begin{enumerate}
\item \emph{confluence of switching}: when transforming derivation
  sequences into equivalent ones by switching independent steps, the
  result not depends on the order in which switchings are applied, but
  only on the resulting permutation;

Unfortunately, we show that for left-linear rewriting systems the characterisation of
shift equivalence based on consistent permutations fails: the
existence of a consistent permutation is only necessary but not
sufficient, hence it cannot be used for proving (1) and (2). 
%
Nevertheless, we will show that property (1) still holds, while only a weak form of (2) does.

\todo[inline]{
DA RIUSARE\\
Firstly, Church-Rosser theorem does not work out of the box. In
particular, sequential independence between rewriting steps~\cite{}
expressed in terms of the existence of a so-called independence pair
is a canonical and well-understood notion of independence in the
linear case. \todo{P: Too technical? Forse qui e' piu' esplicativo un
  esempio} Whenever two steps are sequential independent
\begin{itemize}
\item they can be switched and 
\item the result of the switch is uniquely determined.
\end{itemize}

Here we observe that, in general, both property can fail: sequential
independent steps could be not switchable and different independent
pairs could exist leading to different switches of the same rewriting
steps.  A refined notion of independence needs to be considered to ensure
that independent rewriting steps can be switched, which is based in the concept
of fillers. The problem was identified
in~\cite{baldan2011adhesivity}, where a subclass of adhesive
categories was identified where independence as induced by a filler
coincides with sequential independence (as induced by an independence
pair). Here it is extended to $\mathcal{M}$-adhesive categories.
}

First of all, we will provide a refined notion of independence, which is based 
in the concept of fillers. The issues was identified
in~\cite{baldan2011adhesivity}, where a subclass of adhesive
categories was identified where independence as induced by a filler
coincides with sequential independence (as induced by an independence
pair). Here it is extended to $\mathcal{M}$-adhesive categories.

Then, we move to restrict the class of admissible left-linear rewriting systems,
always working at the level of $\mathcal{M}$-adhesive categories.
Unfortunately, the most natural idea, put forward in~\cite{baldan2017domains}, 
of limiting sequential independence to those cases in which the switch is possible
and uniquely determined does fall short.
%
In this work we focus on what we call
\emph{strongly tame rewriting systems}, where sequential independence ensures existence and
uniqueness of the switch and argue that this is the right setting for
dealing with left-linear rewriting systems.

@@

On the other hand, we show that an appropriate theory of rewriting can
be developed for very tame rewriting systems, recovering, sometimes in
weakened form, results which hold in the linear case.  Local
Church-Rosser is fully recovered for very tame rewriting systems.
